---
title: "CLUSTER ANALYSIS"
author: "Anshy Vislao & Ricardo Flores"
format: html
editor: visual
echo: true
warning: false
---

# PARTITIONING CLUSTERING

## Partitioning Around Medoids

[*(Kaufman & Rousseeuw, 1990)*]{.smallcaps}

We need to install and call the following packages

```{r}
library(tidyverse)
library(sf)
library(factoextra)
library(cluster)
library(NbClust)
library(corrplot)

```

To add information we need to prepare a data frame where columns as variables and rows as elements.

```{r}

rm(list = ls())
data<-read.csv(file.choose())
#data<-read.csv("H:/CLUSTER/estaciones2.csv")
rownames(data) <- c("Andahua","Aplao","Ayo","Cabanaconde","Choco","Chuquibamba", "Crucero_Alto","Huambo","Madrigal","Machahuay","Orcopampa","Pampa_Majes", "Pañe","Porpera","Sibayo","Tisco","Visuyo","Camana","Cotahuasi","Huanca", "Imata","Jana_Cayll","Lagunillas","Joya","Pillones","Puica","Salamanca", "Yanaquihua","Yanque")

```

Then we scale the data to uniform data with different units.

```{r}
stzn.data <- scale(data)
```

We can calculate the distance or dissimilarity matrix in order to use it as an input of the PAM function. With the function `dist()` we can define the observation data and the method ("euclidean","manhattan"). Then, the function PAM could be run as follows

```{r}
dist.data <- dist(stzn.data, method = "manhattan")
```

> The use of seed allows us run again the code with the same results

```{r}
set.seed(2000)
clus.pam <- pam(stzn.data, 4, diss = F, metric = "manhattan")
```

Plotting the silhouette index and calculating the average value for each cluster

```{r}
fviz_silhouette(clus.pam) + theme_bw() + coord_flip()
sltte.plt <- factoextra::fviz_silhouette(
  clus.pam, label = FALSE, print.summary = TRUE
) + theme_bw() + coord_flip() +
  theme(
    axis.text.x = element_text(size = 16),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title = element_text(size = 1)
)
```

Plotting the cluster with the tow main components

```{r}
factoextra::fviz_cluster(
  clus.pam,
  ggtheme = theme_bw(),
  ellipse.type = "t",
  labelsize=12,
  geom = "point",
  star.plot = TRUE
)
```

# HIERARCHICAL CLUSTERING

## AGGLOMERATIVE HIERARCHICAL CLUSTERING

Considering the same data as the first method, we use the function `hclust()` as follow:

```{r}
clus.ahm <- hclust(d=dist.data,method = "ward.D2")
```

<!--# ward.D tiene buen agrupamiento cophenetic: 0.5656147 -->

<!--# ward.D2 cophnetic:  -->

Then, we can visualize the dendogram and calculate the correlation between cophenetic distance and the original distance

```{r}
fviz_dend(clus.ahm,cex=0.5,k=5,k_colors=c("#2E9FDF","#00EEEE","#E7B800","#FC4E07","#7FFF00"))
coph <- cophenetic(clus.ahm)
cor(dist.data, coph)

grp<-cutree(clus.ahm,k=5)
fviz_cluster(list(data=stzn.data,cluster=grp), palette=c("#2E9FDF","#00EEEE","#E7B800","#FC4E07","#7FFF00"))

```

Another alternative is the use of `agnes()` function

```{r}
clus_ag <- agnes(x=dist.data, diss= T, stand = F, metric = "manhattan", method = "ward")
fviz_dend(clus_ag,cex=0.5,k=5,k_colors=c("#2E9FDF","#00AFBC","#E7B800","#FC4E07"))
```
